Structuring College Access: The Market Segment Model and College Board Geo-markets

### 2024-10-18 text/front_end_v2.qmd
revising front end

```{=tex}
\pagenumbering{roman}
\newpage
\pagenumbering{gobble}
\pagenumbering{arabic}
```

@@

In 1983, Zemsky and Oedel authored *The Structure of College Choice*, published by *The College Board.* This book develops the Market Segment Model and introduces local "community-based enrollment markets" (p.14) -- later called "geomarkets" -- which are the basis for the College Board market Enrollment Planning Service (EPS), a market research product. @RN4982[pp., 9-10] write that the Market Segment Model sought to quantify the knowledge that college admissions officers have about student demand:

> Admissions officers invariably are tellers of stories -- about the colleges they represent, about the colleges they attended, about each other, and about the often vagabond life of college recruiting\ldots We believe that the institutions of admissions officers actually comprise a remarkably systematic body of knowledge about the college selection process\ldots Our research, thus, is based on listening carefully to what admissions officers have to say."

@RN4982 [p. 10] stress the importance of "knowing the territory":

@@

The Market Segment Model was created by analyzing College Board data about which colleges SAT test-takers from 1980 sent scores to. High school students who sent scores mostly to colleges in their local geomarket were defined as "local" students. "In-state" students were those who sent most scores to colleges outside their local market but in their home state. "Regional" students were those who sent most scores to colleges outside their home state but in their region (e.g., New England, Midwest). "National" students were those who sent most scores to colleges outside their region. Having categorized students, each local geomarket could be evaluated by the number of students from each of the four market segments, yielding the Market Segment Profile (see @tbl-market-segment-characteristics). Next, for each local geomarket, the Institutional Profile (see @tbl-sample-institutional-profile) the number of students from each market segment that sent scores to a particular college, which could be your college or a competitor.

Collectively, the quantitative outputs of the Market Segment Model provided intelligence about where are the desirable students and which students might consider your college or a competitor. The Market Segment model and geomarkets became the basis or the College Board Enrollment Planning Service (EPS), founded in 1984 and still active today. @RN5019_2010 says "Enrollment Planning Service (EPS\textsuperscript{\textregistered}) the analysis tool that pinpoints the schools and geomarkets where your best prospects are most likely to be found." @RN5021 reports that in 1995, 37% of 4-year publics and 49% of 4-year privates used EPS, while 41% of 4-year publics and 16% of 4-year privates used ACT's market analysis service product, which was similar to EPS.

In 19XX, geomarkets were incorporated into the College Board Student Search Service student list product, which College Board created in 1972 [@belkin2019-studata]. Student lists contain the contact information of prospective students, the basic input for recruiting campaigns that target individual prospects using mail, email, and targeted social media. RN4895 reported that 87% of private and 86% of public four-year institutions purchase student lists. The Student Search Service database consists of College Board test-takers. Colleges pay a fee for each prospect (e.g., \$0.50 in 20XX) and control which prospect profiles through purchase by selecting search filters, such as high school graduation year, SAT score, AP test score, state, etc. Geomarket filters enable colleges to include/exclude prospects from particular geomarkets.

Scholarship in the sociology of education describes schools and colleges as "sorting machines" that reproduce inequality by incorporating, creating, and allocating students to different categories [@RN4873]. Given the consequences subsequent outcomes, the sociology of education is particularly concerned with the transition from high school to college. Over the last decade or so, scholarship on college access has begun to take more seriously the idea that colleges are not passive recipients of applications; rather they expend substantial resources soliciting demand from desirable applicants [@RN3519; @RN4774; @RN4324]. Most scholarship on recruiting analyzes the behavior of individual colleges, thereby assuming that recruiting behavior is a function of individual colleges [e.g., @RN4758; @RN4759]. @RN5012 analyze the College Board *Student Search Service* product, highlighting the role that third-party products play in structuring the recruiting behavior of colleges. The authors show that the utilization of particular search filters -- particularly geographic filters, such as zip code -- is associated with systematic racial exclusion that does not depend on which college purchased the list.

However, prior research has not examined the role of third-party market research in reproducing socioeconomic and racial inequality. This manuscript analyzes the Market Segment Model and College Board Geomarkets as a case study of quantification. We develop arguments from the broader literature on quantification [e.g.,] [@RN5003; @RN5005; @RN5007], particularly the discussions of correlation and homophily from @RN4975. Predictive analytics are based on correlation. When we use data on past correlations to make recommendations about the future, we amplify the effects of historic structural inequality. The Market Segment Model analyzes SAT score-sending data from 1980 and concludes that student demand for higher education is primarily a function of social class= [@RN4982]. Homophily is the idea that actors that share common characteristics are likely to form connections with one another. The Market Segment Model argues that homophily is the organizing principle of student demand and competition between colleges, that "the hierarchical structure of collegiate competition largely reflects the stratified social and economic dimensions of the communities from which colleges draw their students" [@RN4982, p. 72]. Subsequently, the assumption of homophily is programmed into the EPS and Student Service Service products that colleges utilize to decide where to focus recruiting energy.

Empirically, our analyses address the following research questions, which speak to how geomarkets are utilized within the Enrollment Planning Service and within the Student Search Service student list product. First, what is the socioeconomic and racial variation between geomarkets in metropolitan areas and how does this variation change over time? We address this question by spatially joining the geomarket shapefile to Census data about socioeconomic and racial characteristics. Second, how does the socioeconomic and racial composition of included versus included prospects vary when student list purchases filter on particular geomarkets, in concert with other commonly used search filters? We address this question by analyzing student list purchases that do not filter geomarkets and simulating which prospects would have been included had they filtered on particular geomarkets. We obtained these data by issuing public records requests to public universities in four states. Third, in student list purchases that filtered on geomarkets, how does the socioeconomic and racial composition of included students compare the that of the geomarket and how does it compare to that of the broader metropolitan area?

Scholarship on quantification in education has focused on consumer-facing measures such as U.S. News Rankings [@RN5003] and UK "league tables" about school quality [@RN5005]. To our knowledge, prior knowledge has not examined business-facing measures produced by third-party market research. Here, the desirable attributes of students and places they live are quantified with the purpose of informing colleges about where they should allocate recruiting resources to efficiently enroll the most desirable students. We reason that market research reproduces inequality by encouraging colleges to plan their recruiting efforts based on a snapshot of social stratification in student demand that is substantially a function of historic, structural inequality in educational opportunity. In contrast to @RN5005, we cannot demonstrate the effect of geomarkets on inequality. However, we show the mechanisms of how geomarkets reproduce inequality by simulating who is included/excluded when geomarkets are utilized within student list purchases as recommended by @RN4982.

In the following section we provide background information about enrollment management and scholarship on recruiting. Second, we introduce core ideas and concepts from scholarship on quantification. Third, we apply these ideas to a close read of @RN4982 and motivate hypotheses. Four, we describe data and methods. Fifth, we present results. Finally, we discuss implications for scholarship, policy, and practice.

# Background: Enrollment Management

The term "enrollment management" (EM) refers to a profession, an administrative structure, and an industry. As a profession, EM integrates techniques from marketing and economics in order to "influence the characteristics and the size of enrolled student bodies" [@RN2771, p. xiv]. As an administrative structure, the EM office typically controls the activities of admissions, financial aid, and recruiting [@RN2406]. The EM industry consists of university personnel (e.g., admissions counselors, VP for enrollment management), professional associations (e.g., National Association for College Admission Counseling), and third-party servicers that produce market research, software, and and consulting services (e.g., College Board, ACT, EAB, Huron, Slate, PowerSchool).

Figure @fig-em-funnel depicts the "enrollment funnel," which modifies the traditional "marketing funnel" to depict broad stages in the process of recruiting students [@RN4986; @RN2004]. The funnel begins with a large pool of "prospects" (i.e., prospective students) that the university would like to enroll. "Leads" are prospects whose contact information has been obtained. "Inquiries" are prospects that contact the institution and consist of two types: first, inquiries who respond to an initial solicitation (e.g., email) from the university; and second, "student-as-first-contact" inquiries who reach out to the university on their own (e.g., sending ACT scores). <!--  Applicants consist of inquiries who apply plus "stealth applicants" who do not contact the university before applying.  -->The funnel narrows at each successive stage in order to convey the assumption of "melt" (e.g., a subset of "inquiries" will apply). Practically, the purpose of the enrollment funnel is to inform recruiting interventions that target one or more stages. These interventions seek to increase the probability of "conversion" across stages [@RN4322]. At the top of the enrollment funnel, purchasing student lists is the primary means of converting prospects to leads [@list_biz]. Purchased leads are served emails, brochures, and targeted social media advertisements designed to solicit inquiries and applications [@RN4895].

Scholarship at the nexus of enrollment management and college access can be categorized by which part(s) of the enrollment funnel they focus on. The majority of scholarship focuses on the admissions stage, analyzing which admissions criteria are utilized and/or which applicants are admitted [e.g., @RN4874; @RN4993; @RN4131; @RN3522; @poon2022rethinking]. A growing literature analyzes the earlier "recruiting" stages of identifying leads, soliciting inquiries, and soliciting applications. Scholarship on recruiting has conceptualized recruiting behavior as an indicator of college enrollment priorities [@RN4758] and identifies recruiting practices as a mechanism for social reproduction in college access [@RN4324; @RN3519].

Ethnographies by @RN3519 and by @RN4407 analyze the connections between privileged colleges and privileged high schools. @RN4407 shows how guidance counselors at elite private school guidance counselors lobby admissions counselors on behalf of marginal students. @RN3519 finds that college admissions officers visit high schools as a means of maintaining positive relationships with guidance counselors at affluent feeder schools. Consistent with this finding, quantitative case studies of off-campus recruiting visits show that selective private colleges visit affluent, predominantly white schools and communities, with a disproportionate number of visits to private high schools [@RN5014; @RN4330]. Additionally, public research universities often make make more out-of-state recruiting visits than in-state visits, and these out-of-state visits focus on the same set sorts of schools targeted by selective private universities [@RN4759; @RN4758].

Other studies investigate recruiting behavior by non-selective institutions [@RN4520; @cottom2017lower; @dache2018dangling]. Analyzing recruiting from the perspective of high school students, @RN4324 finds that underrepresented students were drawn to colleges that made them feel wanted, often attending institutions with lower graduation rates and requiring larger loans than other college options. @cottom2017lower shows that for-profit colleges found a niche in Black and Latinx communities because traditional colleges ignored these communities. <!--  WHAT YOU SAID ABOUT POSECZNICK Posecznick (2017) studied the “Ravenwood College,” a private nonprofit college founded to serve working Black women. Consistent with analyses of tuition discounting by Lapovsky (2001), the survival of Ravenwood depends on the success of admissions officers in identifying prospects (e.g., through on- and off-campus recruiting events) and “converting” these prospects to enrolled students.  -->

We make two observations about the recruiting literature. First, consistent with scholarship on political economy [@RN4517; @cottom2017lower; @RN5015] and industrial organization [@RN1549; @RN2133], the industry is characterized by a vertical hierarchy of postsecondary institutions matching to a vertical hierarchy of customer niches, a process that exemplifies the idea that higher education is mechanism of social reproduction [@RN2377; @RN4873].<!--  For example, for-profit institutions target adults in communities not being targeted by traditional colleges [@cottom2017lower]. Non-selective private non-profits target prospects who might be considering community college or a local state colleges [@RN4324]. Selective private colleges target private high schools, with a particular link between sectarian high schools and colleges of the same faith [@RN5014], but these schools are increasingly targeted by out-of-state public research universities, suggesting growing competition for a coveted market niche. --> Second, by focusing on the behaviors of individual colleges, the literature implicitly states that recruiting is something done by individual colleges. 
<!--  
Why then, does recruiting behavior look so similar across colleges of a particular "type"?[^1]
[^1]: New institutional theory offers one explanation [@RN527]. In highly developed organizational fields like higher education, forces of competitive and institutional isomorphism -- including institutional locics -- compel organizations adopt similar practices as their peers. Practices deemed advantageous spread quickly, for example, public research universities targeting nonresident prospects to compensate for declines in state appropriations. These behaviors are conceived as autonomous decisions by individual colleges, albeit structured by institutional logics.
 -->

<!-- make or buy  
The process of recruiting students involves many "make or buy" decisions about whether to perform a given task in-house or outsource it to a third-party vendor. EM consulting firms provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest enrollment consulting firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm]. In our data collection, attempting to obtain data about student list purchases from all public universities in four states, at least half of these universities outsourced student list purchases to an EM consultancy [@list_biz; @list_empirics].
-->

However, the recruiting behavior of colleges is structured by third-party actors. Drawing from scholarship on organizational theory [@RN4738], enrollment management processes involve many "make or buy" decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. EM consulting firms provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success. <!-- As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. --> The two largest enrollment consulting firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm].

Moreover, process of college access is structured by third-party recruiting products utilized by colleges. @RN5012 deconstruct the College Board *Student Search Service* product. Selection devices are procedures or routines for making categorical decisions based on input values [@RN4778]. Student list products are selection devices that enable colleges to choose which prospective students they target or ignore based on search filters that incorporate search filters (e.g., high school graduating class, state) which enable colleges to select prospective students. "Racialized inputs" are ostensibly race-neutral inputs that are systematically correlated with race because marginalized racial/ethnic groups have historically been excluded from the input [@RN4786]. @RN5012 argue that several frequently utilized student list filters (zip code, AP test score, SAT score) are racialized inputs. Using a national sample of high school students, and also using data from actual student lists purchased by public universities, they show that racialized search filters have a strong negative relationship with the selection of Black and Latinx prospects. 
 <!-- Using data from actual student lists purchased by public universities, they show that raciaized search filters -- used in conjunction with other search filters -- yield dramatic racial inequality in which prospective students are targeted. -->

```{=html}
<!--The enrollment management industry emphasizes market research.
Market research is the process of producing information that informs decisions about product design or how to find customers.   -->
```

The enrollment management industry puts great stock in market research. @RN4986[p. 17] define market research as all "activities intended to increase understanding of a market and consumer behavior in that market." At the most basic level, in-house market research is a function of being intentional about record-keeping and using those records [@RN3519].^[@RN3519 describes the archives of annual reports of admissions activities from a selective liberal arts college. In the 1942 report, "Section 2.5 is titled 'Secondary School Quartile Standings," a measure of where matriculants fell in the rankings of their high school classes" [@RN3519] while Section 3.11 describes the high schools visited. In the chapter entitled "Travel," @RN3519 [p. 69] describes how The College used record-keeping to inform decisions about visits to local high schools in the 2000s: "The College keeps files on hundreds of schools across the country, and around the globe, with which it had contact \ldots The files enable officers to quickly discern"how we've done here lately," \ldots with the number, relative quality, and yield rate of applicants from a particular school."] However, most market research in the contemporary enrollment management industry is produced by third-party organizations. For example, Ruffalo-Noel Levitz and EAB both produce prodigious amounts of market research, some publicly available as a means of business development [e.g., @RN4741; @RN4728, but most only available to clients. EPS remains the market research arm of College Board, while Encoura Eduventures is the market research arm of ACT.

Despite being central to enrollment management for at least 40 years [@RN4986], the role of third-party market research has been ignored by scholarship on college access. Scholarship on college access often takes a demand-side perspective, whereby household differences in economic and cultural capital yield stratification in student preferences and the ability to gain admission [e.g., @RN4935;@RN1814]. However, market research informs where colleges look for students and which prospective students to target. These recruiting interventions intervene on student demand [@RN4324; @cottom2017lower], thereby structuring college access pathways in ways that could amplify or diminish historic inequality in college access. The College Board played a pivotal role in transforming market research about enrollment from something done by individual universities to something done by third-party professionals [@RN4986]. This article offers one story about the quiet revolution that transformed college access into a sophisticated, supply-led enterprise. We argue that market research structures who colleges recruit, in ways that reinforce existing social stratification. The case of Market Segment Model and College Board geomarkets shows how market research -- and its commodification into recruiting products -- is a mechanism for social reproduction. 
[LAST TWO SENTENCES ARE REDUNDANT BUT BUT ARE TRYING TO DO SLIGHTLY DIFFERENT THINGS AND I LIKE THEM BOTH]

# Quantification

```{=html}
<!--
CUT GENERAL PARAGRAPH ON REPRODUCTION

Within the sociology of education, @RN2377 describes education as a sieve, allocating people to opportunities in socially stratified, capitalist societies. This role highlights "Max Weber's core insight that education has a dual character" [@RN2377, p. 129], on one hand, serving social mobility by redistributing access to privileged positions and, on the other, serving social reproduction by granting privileged families disproportionate access to the educational institutions that confer pathways to privileged positions. Similarly, @RN4873 describe schools and colleges as "sorting machines" that incorporate external categories (e.g., race, class), create categories (e.g., instructional tracks), and allocate students to categories. Initial category allocation affects subsequent allocation, such that school assignments and instructional track assignments affect college access, which affects employment assignments. "In evaluating schools as inequality regimes" @RN4873 \[p. 322\] argue, "the relevant question is not whether schools produce inequality; rather, the relevant questions are, when and where do schools produce inequality, \[and\] how do they do so." As @RN2377 show, whereas early research on social stratification sought to identify patterns of inequality [e.g., @RN4873; @RN290; @RN1268], the sociology of education is now more concerned with identifying the mechanisms underlying these patterns. 
-->
```
We situate our study amidst scholarship on within scholarship on quantification as a mechanism that reproduces inequality. @RN5007 [p. 402] define quantification as "the production and communication of numbers" (p. 402) and introduce reacivity, discipline, and authority as three interrelated themes that describe the effects of quantification. Reactivity is the idea that salient quantitative measures cause people and organizations to change their behavior. Discipline is the idea that quantification causes actors to react in a particular way. Authority is the idea that quantification changes decision-making power, often weakening the discretion of local decision-makers.

>>>>

\pagenumbering{roman}
\newpage
\pagenumbering{gobble}
\pagenumbering{arabic}

@@

In 1983, Robert Zemsky and Penney Oedel authored *The Structure of College Choice*. Based on an analysis of 1980 SAT score-sending behavior and published by *The College Board*, @RN4982 developed the Market Segment Model to help colleges recruit students. The underlying idea of the Market Segment Model is that student demand for your institution is fundamentally a function of class. Therefore, the key to recruitment is identifying territories where large numbers of your class live. @RN4982[p. 11] describe the Market Segment Model as an effort to "quantify" the knowledge that college admissions officers have about the geography of student demand:

@@

@RN4982 also created local "community based markets" (p. 14) -- later known as "Geomarkets." Geomarkets are geographic borders that disaggregate states and large metropolitan areas for recruiting purposes. @fig-nyc-geomarkets shows New York City area Geomarkets. In defining Geomarkets, @RN4982 did not draw new geographic borders, but rather made certain extant borders more salient to the process of recruiting.

Geomarkets have been incorporated into the supply-side structure of college access in ways that are simultaneusly mundane and fundamental. We posit three ways. First, Geomarkets became a principle for how many -- if -- admissions offices organized their labor. In states where a college recruits heavily from, admissions officers are often assigned specific Geomarkets as their territory [CITE]. Geomarkets were created to mirror the territories of admissions officers and, in turn, admissions offices structured their territories around Geomarkets. Second, The Market Segment Model and Geomarkets became the basis for Geomarkets are the basis for the College Board Enrollment Planning Service (EPS), founded in 1984 and still active today. EPS software recommends which Geomarkets a college should recruit from and which schools/communities they should prioritize within targeted Geomarkets. @RN5019_2010 says "Enrollment Planning Service (EPS\textsuperscript{\textregistered}) the analysis tool that pinpoints the schools and geomarkets where your best prospects are most likely to be found." @RN5021 reports that, in 1995, 37% of 4-year public institutions and 49% of 4-year private institutions used EPS.

Third, Geomarkets were incorporated into the College Board "student list" product named Student Search Service. Student lists have been the primary source of lead generation in U.S. higher education since 1972, when College Board began selling "names" [@belkin2019-studata; @list_biz]. @RN4895 reported that 87% of private and 86% of public four-year institutions purchase student lists. Lists contain contact information for prospective students. The Student Search Service database consists of College Board test-takers. Colleges pay a fee for each prospect (e.g., \$0.50 in 2021). Colleges control which prospect profiles they purchase by selecting search filters, such as high school graduation year, SAT score, AP score, state, etc. Geomarket filters enable colleges to include/exclude prospects from particular Geomarkets. An analysis of 830 student lists purchased by 14 public universities found that 16% of purchases filtered on Geomarket [@RN5012].

Geomarkets have never been the subject of empirical research. This omission is surprising because the sociology of education is concerned with sorting [@RN4873], particularly sorting into college. Traditionally, most scholarship -- and certainly most scholarship in economics [HOXBY? CHETTY] -- views inequality in college access as a function households or k-12 schools, which represent the demand-side of higher education. More recently, scholarship in the sociology of education takes seriously the idea that colleges are not passive recipients of applications, but rather they expend substantial resources soliciting demand from desirable applicants [@RN3519; @RN4774; @RN4324]. However, most scholarship on recruiting analyzes the behavior of individual colleges [e.g., @RN4758; @RN4759], thereby creating the assumption that recruiting is a function of individual colleges. What these studies have missed is the broader enrollment management industry -- third-party vendors, their products and consulting solutions -- as a set of mechanisms that structure college access, funneling certain kinds of students to certain kinds of institutions.

This manuscript analyzes one of those mechanisms, The Market Segment Model and College Board Geomarkets, as a case study of quantification. We develop arguments from the broader literature on quantification \[e.g.,\] [@RN5003; @RN5005; @RN5007], particularly the discussions of correlation and homophily from @RN4975. Predictive analytics are based on correlation. When we use data on past correlations to make recommendations about the future, we amplify the effects of historic structural inequality. The Market Segment Model analyzes SAT score-sending data from 1980 and concludes that student demand for higher education is primarily a function of social class [@RN4982]. The Market Segment Model argues that homophily -- actors that share characteristics are likely to form connections --  is the organizing principle of student demand and competition between colleges. @RN4982[p. 72] state that "the hierarchical structure of collegiate competition largely reflects the stratified social and economic dimensions of the communities from which colleges draw their students."  This snapshot of student demand in 1980, itself a consequence historic structural inequality, was programmed into the way admissions offices organize labor and programmed into the EPS and Student Service Service products that colleges utilize to identify and target prospective students. The result is a supply-side that reinforces structural inequalities observed on the demand-side.

Our analyses address the two research questions, which speak to how geomarkets are utilized within EPS and within the Student Search Service student list product. First, what is the socioeconomic and racial variation between geomarkets in metropolitan areas and how does this variation change over time? We address this question by spatially joining the geomarket shapefile to Census data about socioeconomic and racial characteristics. Second, how does the socioeconomic and racial composition of included versus included prospects vary when student list purchases filter on particular geomarkets? We address this question by analyzing actual student list purchases that utilize commonly used search filters (e.g., PSAT score, GPA) but do not filter on Geomarkets. We simulate which prospects in particular metropolitan areas would have been included/excluded had the student list purchase filtered on particular Geomarkets. We obtained these data by issuing public records requests to public universities.
<!-- Third, in student list purchases that filtered on geomarkets, how does the socioeconomic and racial composition of included students compare the that of the geomarket and how does it compare to that of the broader metropolitan area?  -->

<!-- 
THIS TEXT IS FINE, BUT TRYING TO SET UP THE CONTRIBUTION TO BROADER QUANTIFICATION LITERATURE, VIS-A-VIS NO-ONE HAS STUDIED MARKET RESEARCH; DON'T THINK THIS IS NECESSARY AND YOU ARE NOT ON FIRM GROUND HERE IN CALLING THIS "MARKET RESEARCH"
Scholarship on quantification in education has focused on consumer-facing measures such as U.S. News Rankings [@RN5003] and UK "league tables" about school quality [@RN5005]. To our knowledge, prior knowledge has not examined business-facing measures produced by third-party market research. Here, the desirable attributes of students and places they live are quantified with the purpose of informing colleges about where they should allocate recruiting resources to efficiently enroll the most desirable students. We reason that market research reproduces inequality by encouraging colleges to plan their recruiting efforts based on a snapshot of social stratification in student demand that is substantially a function of historic, structural inequality in educational opportunity. In contrast to @RN5005, we cannot demonstrate the effect of geomarkets on inequality. However, we show the mechanisms of how geomarkets reproduce inequality by simulating who is included/excluded when geomarkets are utilized within student list purchases as recommended by @RN4982.
-->

In the following section we provide background information about enrollment management and scholarship on recruiting. Second, we introduce core ideas and concepts from scholarship on quantification. Third, we apply these ideas to a close read of @RN4982 and motivate hypotheses. Four, we describe data and methods. Fifth, we present results. Finally, we discuss implications for scholarship and for policy.

# Enrollment Management and Quantification

**Enrollment Management**

As a profession, enrollment management (EM) integrates techniques from marketing and economics in order to "influence the characteristics and the size of enrolled student bodies" [@RN2771, p. xiv]. As an administrative structure, the EM office typically controls the activities of admissions, financial aid, and recruiting [@RN2406]. @fig-em-funnel depicts the "enrollment funnel," which modifies the traditional "marketing funnel" to depict broad stages in the process of recruiting students [@RN4986; @RN2004]. The funnel begins with a large pool of "prospects" (i.e., prospective students) that the university would like to enroll. "Leads" are prospects whose contact information has been obtained. "Inquiries" are prospects that contact the institution, consisting of those that respond to an initial soliciation (e.g., email) and those that reach out on their own (e.g., sending SAT scores). <!--  Applicants consist of inquiries who apply plus "stealth applicants" who do not contact the university before applying. The funnel narrows at each successive stage in order to convey the assumption of "melt" (e.g., a subset of "inquiries" will apply).  -->The purpose of the enrollment funnel is to inform recruiting interventions that target one or more stages. These interventions seek to increase the probability of "conversion" across stages [@RN4322]. At the top of the enrollment funnel, purchasing student lists is the primary means of converting prospects to leads [@list_biz]. Purchased leads are served emails, brochures, and targeted social media advertisements designed to solicit inquiries and applications [@RN4895].

Scholarship at the nexus of enrollment management and college access can be categorized by which part(s) of the enrollment funnel it speaks to. The majority of scholarship focuses on the admissions stage, analyzing which admissions criteria are utilized and/or which applicants are admitted [e.g., @RN4874; @RN4993; @RN4131; @RN3522; @poon2022rethinking]. Economists often investigate financial aid leveraging, which seeks to convert admits to enrolled students [CITE]. 
<!--  Scholarship on college access often takes a demand-side perspective, whereby household differences in economic and cultural capital yield stratification in student preferences and the ability to gain admission [e.g., @RN4935; @RN1814].   -->

A growing college access literature in sociology analyzes the earlier "recruiting" stages of identifying leads, soliciting inquiries, and soliciting applications. @RN4758 conceptualize recruiting behavior as an indicator of college enrollment priorities. Ethnographies by @RN3519 and @RN4407 identify connections between private school guidance counselors and college admissions officers as a mechanism for social reproduction. Quantitative case-studies of off-campus recruiting visits by public research universities and by selective private universities reveal a preference for visiting private schools and affluent, predominantly white public schools [@RN4758; @RN4759; @RN5023]. From the student perspective, @RN4324 finds that underrepresented students were drawn to colleges that made them feel wanted, often attending institutions with lower graduation rates and requiring larger loans than other college options. @cottom2017lower shows that for-profit colleges found a niche in Black and Latinx communities because traditional colleges ignored these communities. Consistent with scholarship on political economy [@RN4517; @cottom2017lower; @RN5015], this literature reveals a vertical hierarchy of postsecondary institutions matching to a vertical hierarchy of customer niches segmented by class and race, a process consistent with the notion of that higher education is mechanism of social reproduction [@RN2377; @RN4873].

By focusing on the behaviors of individual colleges, scholarship on college access implicitly states that recruiting -- and enrollment management more broadly -- is something done by individual colleges. This statement raises the question, why does recruiting behavior look so similar across colleges of a particular "type." Institutional theory defines the organizational field as "those organizations that, in the aggregate, constitute a recognized area of institutional life: key suppliers, resource and product consumers, regulatory agencies, and other organizations that produce similar services and products" [@RN527, p. 143]. The broader enrollment management (EM) industry consists of university personnel (e.g., admissions counselors, VP for enrollment management), professional associations (e.g., National Association for College Admission Counseling), and third-party servicers that produce market research, software, and consulting services (e.g., College Board, ACT, EAB, Slate, PowerSchool). We argue college access is structured by third-party servicers and products that interact with direct-providers (colleges). Enrollment management is often described as an opaque industry because obtaining concrete information about EM practices is difficult. This is why the title of @RN5024 is *Lifting the Veil on Enrollment Management*. Although sociologists have hinted at the ways enrollment management contributes to inequality in colege access [e.g., @RN2406], scholarship has failed to make third-party servicers and products the object of empirical analysis.
<!--  
[^1]: New institutional theory offers one explanation [@RN527]. In highly developed organizational fields like higher education, forces of competitive and institutional isomorphism -- including institutional locics -- compel organizations adopt similar practices as their peers. Practices deemed advantageous spread quickly, for example, public research universities targeting nonresident prospects to compensate for declines in state appropriations. These behaviors are conceived as autonomous decisions by individual colleges, albeit structured by institutional logics.
 -->

Drawing from scholarship on organizational theory [@RN4738], enrollment management processes involve many "make or buy" decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. EM consulting firms provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success. More recently, the EM consulting business model is based on software-as-service products (e.g., for example, EAB's [Enroll 360](https://eab.com/solutions/enroll360/) product) [@list_biz]. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest enrollment consulting firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm].
<!-- In our data collection, attempting to obtain data about student list purchases from all public universities in four states, at least half of these universities outsourced student list purchases to an EM consultancy [@list_biz; @list_empirics].
-->

The EM industry puts great stock in market research, in part because the EM profession emerged from marketing and market research [@RN4986; @RN2771]. @RN4986 [p. 17] define market research as all "activities intended to increase understanding of a market and consumer behavior in that market." At the most basic level, in-house market research is a function of being intentional about record-keeping and using those records [@RN3519].[^1] In recruiting, market research informs decisions about which prospective students to target with recruiting interventions.<!-- THIS IDEA HAS TO GO SOMEWHERE: IN THE HOOK AT END OF THIS SECTION. These recruiting interventions intervene on student demand [@RN4324; @cottom2017lower], structuring college access pathways in ways that could amplify or diminish historic inequality in college access.  --> How to acquire market research is another make-or-buy decision. The College Board played a pivotal role in transforming market research about enrollment from something done by individual universities to something done by third-party professionals. In 1983, College Board published *Applying Market Research in College Admissions* [@RN4986], which became the basis for their Admitted Student Questionnaire (ASQ) product, and published *The Structure of College Choice* [@RN4982], which became the basis for the Enrollment Planning Service (EPS) product/division. Today, EPS remains the market research arm of College Board, while Encoura Eduventures is the market research arm of ACT. Market research often distills large amounts of data into simpler indicators. Therefore, the quantification literature provides a useful lens to understand market research.

[^1]: @RN3519 describes the archives of annual reports of admissions activities from a selective liberal arts college. In the 1942 report, "Section 2.5 is titled 'Secondary School Quartile Standings," a measure of where matriculants fell in the rankings of their high school classes" [@RN3519] while Section 3.11 describes the high schools visited. In the chapter entitled "Travel," @RN3519 [p. 69] describes how The College used record-keeping to inform decisions about visits to local high schools in the 2000s: "The College keeps files on hundreds of schools across the country, and around the globe, with which it had contact \ldots The files enable officers to quickly discern"how we've done here lately," \ldots with the number, relative quality, and yield rate of applicants from a particular school."

**Quantification**

@RN5007 [p. 402] define quantification as "the production and communication of numbers" (p. 402). They introduce reacivity, discipline, and authority as three interrelated themes that describe the effects of quantification. Reactivity is the idea that salient quantitative measures cause people and organizations to change their behavior. Discipline is the idea that quantification causes actors to react in a particular way. Authority is the idea that quantification changes decision-making power, often weakening the discretion of local decision-makers.

Scholarship about U.S. News & World Report (USNWR) Law School rankings demonstrate the reactivity, discipline, and authority effects of quantification [@RN5003; @RN2356; @RN2376; @RN2375]. Law school rankings were developed as a means of informing prospective students about the relative quality of different programs. Speaking to *reactivity* effects, rankings had the intended effect of affecting where students apply and enroll [CITE]. However, rankings also affected hiring decisions of firms because clients evaluated firms based on the prestige of law schools attended by firm lawyers. Once law schools realized that "important groups of constituents --- students, faculty, trustees, employers, other media --- were using rankings to make decisions that had large consequences for schools \ldots \[then\] schools felt pressured to take them seriously" [@RN5007, p. 415]. Rankings *disciplined* the behavior of law schools. For example, applicants with characteristics valued by the rankings system (e.g., LSAT scores) became more important for decisions about admissions and merit aid [@RN5003]. Rankings also weakened the *authority* of admissions directors and admissions officers to make independent decisions about which applicants to admit, ideal class size, and curricular offerinags [@RN5003]. <!--
In a process akin to sports analytics in Moneyball \[CITE MICHAEL LEWIS\], we suggest that market research replaces the wisdom and authority of local admissions counselors.    -->

The analysis of UK League Tables by @RN5005 show how quantification can be a mechanism of social reproduction. In 1992, the UK government began publishing school "league tables," which ranked schools based on student performance on high-stakes national exams taken at age 16. League tables facilitate making evaluative comparisons between schools regardless of geographic proximity. They also discipline households and schools to conceive of school quality in terms of test scores, which are substantially a function of the class composition of schools. Using Census data measured at the local authority level (similar to a U.S. county), @RN5005 find that localities with higher performing schools experienced growth in the share of managerial/professional households following the adoption of league tables and a decline in the share of working-class households. Using longitudinal survey data, managerial/professional households were more likely to move to localities with higher ranked schools following the introduction of school league tables. The introduction of league tables contributed to social reproduction because professional/managerial households were more aware of these consumer-facing metrics and had resources to respond by moving to more expensive neighbhorhoods, near higher performing schools.<!-- THIS IS RATCHET EFFECT STUFF. I CUT IT BECAUSE YOU HAVEN'T INTRODUCED THE IDEA OF RATCHET EFFECT
Schools that performed well on initial league tables tended to be schools with a high share of professional/managerial families. Therefore, quantification takes a snapshot of existing social stratification and that snapshot becomes a power unto itself, amplifying the effect of previous strctural inequality on subsequent structural inequality.
-->

<!-- THIS IS THE CUT TRANSITION PARAGRAPH
In the broader quantification literature, critical data studies [e.g., @RN4975; @RN4849; @RN4772] and sociological analyses of technology and inequality [e.g., @RN4778; @RN4786; @RN4877; @RN4810] introduce concepts salient to the analysis of market research. Our focus is on "selection devices," which utilize some combination of inputs and discretion to categorize places, people, and organizations for the purpose of allocating opportunities or interventions [@RN4778; @RN5012].
-->
